---
title: "New York State COVID-19"
author: "Ed Young"
date:  '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    css: ~/repos/styles.css
    highlight: null
    theme: null
  df_print: kable
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
     echo = FALSE,
     warning = FALSE,
     message = FALSE,
     cache = FALSE
)
check_load_pkgs <- function() {
     if (!require(tidyverse)) {
          message("installing the 'tidyverse' package")
          install.packages(tidyverse)
     }
     if (!require(ggrepel)) {
          message("installing the 'ggrepel' package")
          install.packages(ggrepel)
     }
     if (!require(lubridate)) {
          message("installing the 'lubridate' package")
          install.packages(lubridate)
     }
     if (!require(EpiEstim)) {
          message("installing the 'EpiEstim' package")
          install.packages(EpiEstim)
     }
     if (!require(RSocrata)) {
          message("installing the 'RSocrata' package")
          install.packages(RSocrata)
     }
     if (!require(knitr)) {
          message("installing the 'knitr' package")
          install.packages(knitr)
     }
     if (!require(sf)) {
          message("installing the 'sf' package")
          install.packages(sf)
     }
     if (!require(tigris)) {
          message("installing the 'tigris' package")
          install.packages(tigris)
     }
     if (!require(usmap)) {
          message("installing the 'usmap' package")
          install.packages(usmap)
     }
     if (!require(tidycensus)) {
          message("installing the 'tidycensus' package")
          install.packages(tidycensus)
     }
     if (!require(scales)) {
          message("installing the 'scales' package")
          install.packages(scales)
     }
     if (!require(viridis)) {
          message("installing the 'viridis' package")
          install.packages(viridis)
     }
}
check_load_pkgs()
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)

## Download latest NY State data from health.data.ny.gov
## Munge

nys <-
     read.socrata("https://health.data.ny.gov/resource/xdss-u53e.json") %>%
     select(t = test_date,
            I = new_positives,
            C = cumulative_number_of_positives,
            T = total_number_of_tests,
            county) %>%
     mutate(
          t = as.Date(t),
          I = as.integer(I),
          C = as.integer(C),
          T = as.integer(T),
          f = fips("NY", county = county)
     ) %>%
     arrange(f, t)
nysCounty1 <- nys %>%
     group_by(county) %>%
     filter(max(I) > 100) %>%
     ungroup()
nysCounty2 <- nys %>%
     group_by(county) %>%
     filter(max(I) > 10 & max(I) <= 100)
nysCounty3 <- nys %>%
     group_by(county) %>%
     filter(max(I) <= 10)
state <- nys %>%
     group_by(t) %>%
     summarise(SI = sum(I),
               SC = sum(C),
               ST = sum(T))

## Download latest national data at State level from New York Times
## Munge

Cases <-
     read.csv(
          "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv",
          header = TRUE,
          sep = ","
     ) %>%
     filter(fips < 60) %>%
     mutate(
          date = as.Date(date),
          state = as.character(state),
          f = as.character(as.numeric(fips))
     ) %>%
     arrange(f, date) %>%
     as_tibble()
Cases2 <- Cases %>%
     select(f, date, cases) %>%
     arrange(f, date) %>%
     group_by(f) %>%
     mutate(I = if_else(
          cases - lag(cases, n = 1, default = 0) > 0,
          cases - lag(cases, n = 1, default = 0),
          0
     )) %>%
     ungroup()

## Download data on deaths at County level from NY Times
## Munge

USDeaths <-
     read.csv(
          "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv",
          header = TRUE,
          sep = ","
     ) %>%
     mutate(
          date = as.Date(date),
          county = as.character(county),
          state = as.character(state)
     ) %>%
     as_tibble()
NYDeaths <- USDeaths %>%
     filter(state == "New York") %>%
     select(date, deaths) %>%
     group_by(date) %>%
     summarise(SD = sum(deaths)) %>%
     mutate(d = SD - lag(SD, n = 1, default = 0)) %>%
     ungroup()
NYCountyDeaths <- USDeaths %>%
     filter(state == "New York") %>%
     select(county, date, deaths) %>%
     group_by(county) %>%
     arrange(county, date) %>%
     mutate(d = deaths - dplyr::lag(deaths, n = 1, default = 0)) %>%
     ungroup()
NYCountyDeaths1 <- NYCountyDeaths %>%
     group_by(county) %>%
     filter(max(d) > 100 & county != "Unknown") %>%
     ungroup()
NYCountyDeaths2 <- NYCountyDeaths %>%
     group_by(county) %>%
     filter(max(d) > 10 & max(d) <= 100) %>%
     ungroup()
NYCountyDeaths3 <- NYCountyDeaths %>%
     group_by(county) %>%
     filter(max(d) <= 10) %>%
     ungroup()

## Download US State map

United_States_map <- states() %>%
     filter(STATEFP < 60) %>%
     mutate(f = as.character(as.numeric(STATEFP))) %>%
     select(f, STUSPS)

## Download New York County map data

ny_map <- counties(state = 36) %>%
     mutate(f = paste0(STATEFP, COUNTYFP)) %>%
     select(GEOID, County = NAME, f)


## Download census data
## Munge

NYpop <-
     get_acs(geography = "county",
             state = 36,
             variables = "B01003_001") %>%
     select(GEOID,  estimate)

USpop <-
     get_acs(geography = "state",
             variables = "B01003_001") %>%
     filter(GEOID < 60) %>%
     mutate("f" = as.character(as.numeric(GEOID))) %>%
     select(f,
            "State" = NAME,
            "Population" = estimate)



## Calculate Rt using the EpiEstim Package for New York State
## The window for estimating Rt can be adjusted using t_start and t_end.
## The default setting for the window is 7 days


R.instantaneous <- estimate_R(state$SI,
                              method = "parametric_si",
                              config = list(mean_si = 3.96,
                                            std_si = 4.75))

###### Setup for estimating Rt

## 1. Start with a data.frame that has the following columns: fips, date, incidence
## 2. Obtain the unique list of fips (lof) for which you want to calculate Rt
## 3. Iterate over the list of fips, obtain the list of incidence for each fips
## 4. Use the estimate_R function to estimate R and it's std for each fips
## 5. Assign those values to an ROutput data.frame


makeRoutput <- function(df, tn) {
     lof <-
          df %>% select(f) %>% unique() %>% filter(!is.na(f)) %>% as.list()
     ROutput <- tibble(
          f = lof$f,
          R = 0,
          Std = 0,
          Confl = 0,
          Confh = 0
     )
     for (fips in lof$f) {
          x <- df %>% filter(f == fips, !is.na(I)) %>% select(I) %>% sum()
          y <-
               df %>% filter(f == fips, !is.na(I)) %>% select(I) %>% lengths()
          if (x * 7 / y > 11) {
               t_start <- seq(2, y - tn)
               t_end <- t_start + tn
               R0 <- estimate_R(
                    incid = df %>% filter(f == fips, !is.na(I)) %>% select(I),
                    method = "parametric_si",
                    config = make_config(
                         list(
                              mean_si = 3.96,
                              std_si = 4.75,
                              t_start = t_start,
                              t_end = t_end
                         )
                    )
               )
               n <- length(R0$R$t_start)
               std <- round(R0$R$'Std(R)'[n], 3)
               R <- round(R0$R$'Mean(R)'[n], 3)
               ROutput$R[ROutput$f == fips] <- R
               ROutput$Std[ROutput$f == fips] <- std
               ROutput$Confl[ROutput$f == fips] <-
                    R - 1.96 * std / sqrt(n)
               ROutput$Confh[ROutput$f == fips] <-
                    R + 1.96 * std / sqrt(n)
          } else {
               ROutput$R[ROutput$f == fips] <- NA
               ROutput$Std[ROutput$f == fips] <- NA
               ROutput$Confl[ROutput$f == fips] <- NA
               ROutput$Confh[ROutput$f == fips] <- NA
          }
     }
     return(ROutput)
}


ROutput.NY <- makeRoutput(nys, 7)
ROutput.US <- makeRoutput(Cases2, 7)

#####################################################################################
# Calculate the Effect Size of an intervention by noting the difference between the
# predicted value from the regression line before the intervention, and the value
# from the regression line after the intervention at the same point after the
# intervention.

forNYS <- nysCounty1
forNYS$I[forNYS$I == 0] <- NA
clength <- length(unique(forNYS$county))
CSD <- tibble(
     "County" = unique(forNYS$county),
     b = rep.int(0, clength),
     a = rep.int(0, clength),
     d = rep.int(0, clength),
     g = rep.int(0, clength),
     y_int1 = rep.int(0, clength),
     y_int2 = rep.int(0, clength),
     y_int3 = rep.int(0, clength)
)
for (c in unique(forNYS$county)) {
     x <-
          lm(log10(I) ~ t, data = filter(forNYS, t <= as.Date("2020-03-20") &
                                              county == c), na.action=na.omit)
     y <-
          lm(log10(I) ~ t, data = filter(forNYS, t >= as.Date("2020-04-03") &
                                              county == c), na.action=na.omit)
     z <-
          10 ^ (predict(x, data.frame(t = as.Date("2020-04-03")))) - 10 ^ y$fitted.values[1]
     CSD$b[CSD$County == c] <- x$coefficients[2]
     CSD$a[CSD$County == c] <- y$coefficients[2]
     CSD$d[CSD$County == c] <- x$coefficients[2] - y$coefficients[2]
     CSD$g[CSD$County == c] <- z
     CSD$y_int1[CSD$County == c] <- 10^tail(x$fitted.values, 1)
     CSD$y_int2[CSD$County == c] <- 10^(tail(x$fitted.values, 1)+14*x$coefficients[2])
     CSD$y_int3[CSD$County == c] <- 10^y$model$`log10(I)`[1]
}

CSD <- CSD %>%
     arrange(desc(g)) %>%
     mutate(
          "SlopeBefore" = round(b, 3),
          "SlopeAfter" = round(a, 3),
          "SlopeDiff" = round(d, 3),
          "EffectSize" = round(g, 0)
     ) %>%
     select(
          County,
          "SlopeBefore",
          "SlopeAfter",
          "SlopeDiff",
          "EffectSize",
          y_int1,
          y_int2,
          y_int3
     )

forNYS <- forNYS %>% 
     inner_join(CSD, by = c("county"="County"))

#####################################################
# Joining data from Map, Census, and Rt estimation

ny_map3 <- ny_map %>%
     inner_join(ROutput.NY, by = "f") %>%
     inner_join(NYpop, by = "GEOID")

ny_map4 <- ny_map3 %>%
     as_tibble() %>%
     arrange(County) %>%
     select(County,
            R,
            Confl,
            Confh,
            "Population" = estimate)

us_map <- ROutput.US %>%
     inner_join(United_States_map, by = "f") %>%
     inner_join(USpop, by = "f")

us_map48 <- us_map %>%
     filter(!(
          State %in% c("Alaska", "District of Columbia", "Hawaii", "Puerto Rico")
     ))

us_map4 <- us_map48 %>%
     arrange(State) %>%
     select(State,
            R,
            Confl,
            Confh,
            Population) %>%
     tibble()

###########################################################
# Barrier-free color palette
# Source: Okabe & Ito (2008): Color Universal Design (CUD):
#         Fig. 16 of <https://jfly.uni-koeln.de/color/>:

# (a) Vector of colors (as RGB values):
o_i_colors <- c(
     rgb(0,   0,   0, maxColorValue = 255),
     # black
     rgb(230, 159,   0, maxColorValue = 255),
     # orange
     rgb(86, 180, 233, maxColorValue = 255),
     # skyblue
     rgb(0, 158, 115, maxColorValue = 255),
     # green
     rgb(240, 228,  66, maxColorValue = 255),
     # yellow
     rgb(0, 114, 178, maxColorValue = 255),
     # blue
     rgb(213,  94,   0, maxColorValue = 255),
     # vermillion
     rgb(204, 121, 167, maxColorValue = 255)   # purple
)

# (b) Vector of color names:
o_i_names <-
     c("black",
       "orange",
       "skyblue",
       "green",
       "yellow",
       "blue",
       "vermillion",
       "purple")

names(o_i_colors) <- o_i_names

###########################################################
# Themes

theme(
          plot.background = element_rect(fill = "linen"),
          panel.background = element_rect(fill = "linen"),
          axis.line = element_line(color = "black"),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()
     )
```


<p class="intro">This page is based on data from <a aria-describedby="footnote-label" href="#D1">New York State Department of Health</a>, most recently updated on `r max(nys$t)`; and on data from <a aria-describedby="footnote-label" href="#D2">The New York Times</a>, most recently updated on `r max(NYDeaths$date)`.</p>

The first set of graphs represent new cases of COVID-19 for the state of New York.  Following are graphs and figures representing the same information broken down by county in New York.

Because of the latency from time of infection to symptoms, and then to diagnosis, it may take up to 14 days after a change in policy for the number of new cases to decrease.  I have indicated this two week period after the policy of social distancing enforcement with gray dots.  Regression lines before and after the policy omit that two week period. 

The effect of an intervention can be measured in two ways.  The first, and most intuitive, is to produce two regression lines based on the data before the intervention, and after the intervention.  The first regression line represents the rate of infection before any intervention.  We can extend that line through the two week intervention period, and estimate the number of infections that would have been expected if no intervention had taken place.  Then we can measure the distance from that counterfactual predicted value to the actual number of infections based on the regression line after the intervention.  This gives an estimation of the difference between what the number of infections might have been without an intervention, and the actual number with an intervention - that is, the effect size of the intervention.  A second, more abstract, way is to compare the slopes of the two regression lines.  Although this is a measurement of the intervention, it is not intuitively clear what that change in slope represents.

These charts are for demonstration purposes only.  The regression lines are drawn using simple linear regression, rather than an ARIMA model, which is more appropriate for time series.  In addition, there may be other factors beside the social distancing intervention that may have influenced the change in regression lines during this period.  For example, the natural history of infections may have an increasing phase and decreasing phase without any intervention. However, the fact that we see a marked change at the precise time that social distancing was enforced is strong evidence that it was that intervention that caused the change.

The most recent measurement on each graph is labeled in red.


# New York State Cases

```{r NYS New}
ggplot(data = state) +
  geom_point(mapping = aes(x = t, y = SI)) +
  geom_point(data = filter(state, t == max(t)),
             mapping = aes(x = t, y = SI),
             color = "red") +  
  geom_label_repel(
    data = filter(state, t == max(t)),
    aes(
      x = t,
      y = SI,
      label = sprintf("%i", SI),
      fill="red"
    ),
    show.legend = FALSE,
    segment.color = "black",
    nudge_x = -10,
    nudge_y = 50
  ) +
  geom_vline(xintercept = as.Date("2020-03-20"), linetype = 4) +
  geom_rect(
    aes(
      xmin = as.Date("2020-03-20"),
      xmax = as.Date("2020-04-03"),
      ymin = 0,
      ymax = Inf
    ),
    fill = "gray100",
    alpha = 0.01
  ) +
  geom_label(label = "Social Distancing Enforced",
             x = as.Date("2020-03-20"),
             y = max(state$SI))  +
  labs(x = "Date",
       y = "New Cases per Day") +
  ggtitle(label = "New York State - New Cases", subtitle = "Natural Scale") +
  theme(
    plot.background = element_rect(fill = "linen"),
    panel.background = element_rect(fill = "linen"),
    axis.line = element_line(color = "black"),
    plot.title = element_text(
      face = "bold",
      size = rel(1.5),
      hjust = 0
    ),
    plot.subtitle = element_text()
  )

state1 <- state
state1$SI[state1$SI == 0] <- 1
before <-
  lm(log10(SI) ~ t, data = filter(state1, t <= as.Date("2020-03-20")))
after <-
  lm(log10(SI) ~ t, data = filter(state, t >= as.Date("2020-04-03")))
ggplot(data = state) +
  geom_point(mapping = aes(x = t, y = SI)) +
  geom_point(data = filter(state, t == max(t)),
             mapping = aes(x = t, y = SI),
             color = "red") +  
  geom_smooth(
    data = filter(state, t <= as.Date("2020-03-20")),
    mapping = aes(x = t, y = SI),
    colour = "red",
    method = lm,
    na.rm = TRUE,
    se = FALSE
  ) +
  geom_smooth(
    data = filter(state, t >= as.Date("2020-04-03")),
    mapping = aes(x = t, y = SI),
    colour = "green",
    method = lm,
    na.rm = TRUE,
    se = FALSE
  ) +
  geom_segment(
    aes(
      x = as.Date("2020-03-20"),
      y = 10 ^ before$model$`log10(SI)`[18],
      xend = as.Date("2020-04-03"),
      yend = 10 ^ (before$model$`log10(SI)`[18] + 14 * before$coefficients[2]),
    ),
    linetype = 2,
    color = "red",
    show.legend = NA
  ) +
  geom_segment(
    aes(
      x    = as.Date("2020-04-03"),
      y    = 10 ^ (before$model$`log10(SI)`[18] + 14 * before$coefficients[2]),
      xend = as.Date("2020-04-03"),
      yend = 10 ^ after$model$`log10(SI)`[1]
    ),
    linetype = 2,
    color = "black",
    show.legend = NA
  ) +
  geom_rect(
    aes(
      xmin = as.Date("2020-03-20"),
      xmax = as.Date("2020-04-03"),
      ymin = 0,
      ymax = Inf
    ),
    fill = "gray100",
    alpha = 0.01
  ) +
  geom_label_repel(
    data = filter(state, t == max(t)),
    aes(
      x = t,
      y = SI,
      label = sprintf("%i", SI),
      fill="red"
    ),
    show.legend = FALSE,
    segment.color = "black",
    nudge_x = -10,
    nudge_y = 1
  ) +
  geom_vline(xintercept = as.Date("2020-03-20"), linetype = 4) +
  geom_label(label = "Social Distancing Enforced",
             x = as.Date("2020-03-20"),
             y = 0) +
  geom_text(
    aes(
      label = sprintf(
      "Difference in Slope = %.3f",
      before$coefficients[2] - after$coefficients[2]
    ),
    x = as.Date("2020-04-15"),
    y = 50000
  ), hjust=0
  ) +
  geom_text(
    aes(
      label = sprintf(
        "Effect Size = %.0f",
        10 ^ (before$model$`log10(SI)`[18] + 14 * before$coefficients[2]) - 10 ^
        after$model$`log10(SI)`[1]
      ),
    x = as.Date("2020-04-15"),
    y = 100000
    ), hjust=0
    ) +
  labs(x = "Date",
       y = "New Cases per Day") +
  scale_y_log10(limits = c(1, 2000000), labels = comma) +
  ggtitle(label = "New York State - New Cases", subtitle = "Log Scale") +
  theme(
    plot.background = element_rect(fill = "linen"),
    panel.background = element_rect(fill = "linen"),
    axis.line = element_line(color = "black"),
    plot.title = element_text(
      face = "bold",
      size = rel(1.5),
      hjust = 0
    ),
    plot.subtitle = element_text()
  )
```

\pagebreak

# Cummulative Cases

<p class="intro">The total number of confirmed cases in New York State represents `r round(state$SC[state$t == max(state$t)]/193781, 1)`% of the population.>/p>

```{r NYS Cummulative}
ggplot(data = state) +
  geom_point(mapping = aes(x = t, y = SC)) +
  geom_rect(
    aes(
      xmin = as.Date("2020-03-20"),
      xmax = as.Date("2020-04-03"),
      ymin = 0,
      ymax = Inf
    ),
    fill = "gray100",
    alpha = 0.01
  ) +
  geom_point(data = filter(state, t == max(t)),
             mapping = aes(x = t, y = SC),
             color = "red") +
  geom_label_repel(
    data = filter(state, t == max(t)),
    aes(
      x = t,
      y = SC,
      label = sprintf("%i", SC),
      fill="red"
    ),
    show.legend = FALSE,
    segment.color = "black",
    nudge_x = -10,
    nudge_y = -1
  ) +
  labs(x = "Date",
       y = "Log Scale of Cummulative Cases") +
  scale_y_log10(breaks = c(10, 100, 1000, 10000, 100000),
                labels = comma) +
  geom_vline(xintercept = as.Date("2020-03-20"), linetype = 4) +
  geom_label(label = "Social Distancing Enforced",
             x = as.Date("2020-03-20"),
             y = 1) +
  ggtitle(label = "Cummulative: New York State New Cases", subtitle = "Log Scale") +
  theme(
    plot.background = element_rect(fill = "linen"),
    panel.background = element_rect(fill = "linen"),
    axis.line = element_line(color = "black"),
    plot.title = element_text(
      face = "bold",
      size = rel(1.5),
      hjust = 0
    ),
    plot.subtitle = element_text()
  ) 
```

\pagebreak

# Percentage of Tests that were Positive

<p class="intro">The raw number of positive tests per day can vary depending on how many tests were performed on that day.  One way to normalize this variation is to look at the percentage of tests performed that were positive rather than the raw number of positive tests.</p>

```{r NYS Percentage}
ggplot(data = state) +
  geom_point(mapping = aes(x = t, y = SI * 100 / ST)) +
  geom_point(data = filter(state, t == max(t)),
             mapping = aes(x = t, y = SI*100/ST),
             color = "red") +  
  geom_rect(
    aes(
      xmin = as.Date("2020-03-20"),
      xmax = as.Date("2020-04-03"),
      ymin = 0,
      ymax = Inf
    ),
    fill = "gray100",
    alpha = 0.01
  ) +
  geom_smooth(mapping = aes(x = t, y = SI * 100 / ST),
              method = "gam") +
  geom_label_repel(
    data = filter(state, t == max(t)),
    aes(
      x = t,
      y = SI * 100 / ST,
      label = sprintf("%.1f %%", SI * 100 / ST),
      fill = "red"
    ),
    show.legend = FALSE,
    segment.color = "black",
    nudge_x = -20,
    nudge_y = 50
  ) +
  geom_ribbon(aes(
    ymin = 0,
    ymax = ST / 500,
    x = t,
    fill = ST
  ),
  alpha = 0.3,
  fill = "tomato") +
  scale_y_continuous(sec.axis = sec_axis(~ . * 500, name = "Total Tests Performed")) +
  geom_vline(xintercept = as.Date("2020-03-20"), linetype = 4) +
  geom_label(label = "Social Distancing Enforced",
             x = as.Date("2020-03-20"),
             y = -5) +
  labs(x = "Date",
       y = "Percentage of Tests Performed that were Positive") +
  ggtitle(label = "Precentage Positive: New York State New Tests", subtitle = "Number of Tests Performed") +
  theme(
    axis.title.y.left = element_text(color = "blue"),
    axis.title.y.right = element_text(color = "tomato"),
    axis.text.y.right = element_text(color = "tomato"),
    panel.background = element_rect(fill = "linen"),
    plot.background = element_rect(fill = "linen"),
    axis.line.x = element_line(color = "black"),
    axis.line.y.left = element_line(color = "black"),
    axis.line.y.right = element_line(color = "tomato"),
    plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
    plot.subtitle = element_text(color = "tomato")    
  ) 
```
\pagebreak

# Deaths in New York State

```{r NYS Deaths}
ggplot(data = NYDeaths) +
     geom_point(mapping = aes(x = date, y = d)) +
     geom_point(data = filter(NYDeaths, date == max(date)),
             mapping = aes(x = date, y = d),
             color = "red") +  
     geom_label_repel(
          data = filter(NYDeaths, date == max(date)),
          aes(
               x = date,
               y = d,
               label = sprintf("%i", d),
               fill = "red"
          ),
          show.legend = FALSE,
          segment.color = "black",
          nudge_x = -10,
          nudge_y = -40
     ) +
     geom_smooth(mapping = aes(x = date, y = d),
                 method = "gam") +
     labs(x = "Date",
          y = "Deaths") +
     theme(
          plot.background = element_rect(fill = "linen"),
          panel.background = element_rect(fill = "linen"),
          axis.line = element_line(color = "black"),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()        
     ) +
     ggtitle(label = "Deaths per Day", subtitle = "Natural Scale")

ggplot(data = NYDeaths) +
     geom_point(mapping = aes(x = date, y = SD)) +
     geom_point(data = filter(NYDeaths, date == max(date)),
             mapping = aes(x = date, y = SD),
             color = "red") +
     scale_y_log10() +
     geom_label_repel(
          data = filter(NYDeaths, date == max(date)),
          aes(
               x = date,
               y = SD,
               label = sprintf("%i", SD),
               fill="red"
          ),
          show.legend = FALSE,
          segment.color = "black",
          nudge_x = -10,
          nudge_y = -1
     ) +
     labs(x = "Date",
          y = "Deaths") +
     theme(
          plot.background = element_rect(fill = "linen"),
          panel.background = element_rect(fill = "linen"),
          axis.line = element_line(color = "black"),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()
     ) +
     ggtitle(label = "Cummulative Deaths", subtitle = "Log Scale")
```

# Estimation of R~t~ in New York State as a whole

<p class="intro">Transmissibility of a virus during an epidemic can be measured by the reproduction number R<sub>t</sub>, the average number of secondary infections caused by a primary infected <a aria-describedby="footnote-label" href="#R"> individual</a>.  Statistical methods to estimate R<sub>t</sub> have been developed which require only case incidence data and the distribution of the serial interval (the time between the onset of symptoms in the primary case and the onset of symptoms in the secondary cases). The distribution of the serial interval (mean and standard deviation) can be estimated by directly interviewing people who have been <a aria-describedby="footnote-label" href="#SI">infected</a>.  So long as some new infections continue to occur R<sub>t</sub> will be greater than 0.  If R<sub>t</sub> is less than 1 (each infected individual causes less than one other person to become infected, on average), we would expect the number of new cases each day to be less than the day before.  If R<sub>t</sub> is greater than 1, we would expect the number of new cases to grow.</p> 

The estimate of the most recent R<sub>t</sub> for the State of New York is <b> `r round(R.instantaneous$R$'Mean(R)'[length(R.instantaneous$R$t_start)], 3)` </b>
The 95% Confidence interval around this estimate is quite narrow, and can't be seen clearly given the scale of the graph.  A Histogram of the estimates of R<sub>t</sub> at the most recent point is given below.

```{r NYS R Plot}
estimate_R_plots(
     R.instantaneous,
     what = "R",
     options_R = list(xlab = "Days Since First Reported Infection"),
     legend = TRUE
)

R.sample <-
sample_posterior_R(R.instantaneous,
n = 1000,
window = length(R.instantaneous$R$t_start))
hist(R.sample)
```
\pagebreak

# Number of New Cases by County

<p class="intro">I have divided the counties into three groups:  those that had more than 100 new cases on at least one day; those that had between 10 and 100 cases on one day; and those that had less than 10 cases on any day.</p>

I only ran regressions on those counties with more than 100 cases on a day.

As a reminder, New York City is divided into five boroughs, each borough being its own county.  Manhattan is New York County, the Bronx is Bronx County, Queens is Queens County, Brooklyn is Kings County, and Staten Island is Richmond County. 

```{r County New 1,  fig.height=20, fig.width=20}
ggplot(data = forNYS) +
     geom_point(mapping = aes(x = t, y = I)) +
     geom_point(data = filter(forNYS, t == max(t)),
             mapping = aes(x = t, y = I),
             color = "red") +
     geom_rect(
          aes(
               xmin = as.Date("2020-03-20"),
               xmax = as.Date("2020-04-03"),
               ymin = 0,
               ymax = Inf
          ),
          fill = "gray100",
          alpha = 0.01
     ) +
     labs(x = NULL,
          y = "New Cases per Day") +
     scale_y_log10(labels = comma) +
     geom_vline(xintercept = as.Date("2020-03-20"), linetype = 4) +
     geom_label_repel(
          data = filter(forNYS, t == max(t)),
          aes(
               x = t,
               y = I,
               label = sprintf("%i", I),
               fill="red"
          ),
          show.legend = FALSE,
          segment.color = "black",
          nudge_x = -10,
          nudge_y = 1
     ) +
     geom_smooth(
          data = filter(forNYS, t <= as.Date("2020-03-20")),
          mapping = aes(x = t, y = I),
          colour = "red",
          method = lm,
          na.rm = TRUE,
          se = FALSE
     ) +
     geom_smooth(
          data = filter(forNYS, t >= as.Date("2020-04-03")),
          mapping = aes(x = t, y = I),
          colour = "green",
          method = lm,
          na.rm = TRUE,
          se = FALSE
     ) +
     geom_segment(
          aes(
               x = as.Date("2020-03-20"),
               y = y_int1,
               xend = as.Date("2020-04-03"),
               yend = y_int2
          ),
          linetype = 2,
          color = "red",
          show.legend = NA
     ) +
     geom_segment(
          aes(
               x    = as.Date("2020-04-03"),
               y    = y_int2,
               xend = as.Date("2020-04-03"),
               yend = y_int3
          ),
          linetype = 2,
          color = "black",
          show.legend = NA
     ) +
     geom_text(
          aes(
               label = sprintf("Effect Size is %.0f", EffectSize),
               x = as.Date("2020-04-20"),
               y = 100000
               ),
          hjust=0
          ) +
     geom_text(
          aes(
               label = sprintf("Difference in Slope is %.3f", SlopeDiff),
               x = as.Date("2020-04-20"),
               y = 50000
               ),
          hjust=0
          ) +
     facet_wrap(vars(county), ncol = 4) +
     theme(
          plot.background = element_rect(fill = "linen"),
          panel.background = element_rect(fill = "linen"),
          strip.background = element_rect(fill = "grey80"),
          strip.text = element_text(color = "black"),
          legend.position = "bottom",
          panel.border = element_rect(
               colour = "black",
               fill = NA,
               size = 1
          ),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()         
     ) +
     ggtitle(label="New Cases by County", subtitle = "Counties with a maximum of over 100 New Cases on a single day")
```

# Effect Size

<p class="intro">Note that greatest change in the Effect Size and the slope of the regession lines is in the largest and heaviest hit counties.</p>


```{r County Slope Difference, fig.width=7.5, fig.height=7.5}
#CSD %>%  kable()
ggplot(CSD, aes(y=County, x = `EffectSize`, size = `SlopeDiff`, color = `SlopeDiff`)) +
     geom_point() +
     scale_color_viridis_c() + 
     annotate(geom = "rect", 
           xmin = c(0, 100000, 200000, 300000),
           xmax = c(100000, 200000, 300000, 400000),
           ymin = -Inf,
           ymax = Inf,
           fill = c(o_i_colors[2:5]),
           alpha = 0.1) +
     scale_x_continuous(labels = comma,
                        limits = c(-300, 400000)) +
     guides(color = guide_legend(title = "Difference in Slope"), 
            size = guide_legend(title = "Difference in Slope")
            ) +
     labs(x="Effect Size",
          y=NULL) +
  ggtitle(label = "Social Distancing Effect Size by County", subtitle = "Counties with more than 100 cases in a day") +
     theme(
          panel.background = element_rect(fill = "linen"),
          rect = element_rect(fill = "linen"),
          plot.background = element_rect(fill = "linen"),
          legend.background = element_rect(fill = "linen"),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()        
     )    
```

# New Cases by County

```{r County New 2,  fig.height=30, fig.width=20}
ggplot(data = nysCounty2) +
  geom_point(mapping = aes(x = t, y = I)) +
  geom_point(data = filter(nysCounty2, t == max(t)),
             mapping = aes(x = t, y = I),
             color = "red") +
  geom_rect(
    aes(
      xmin = as.Date("2020-03-20"),
      xmax = as.Date("2020-04-03"),
      ymin = 0,
      ymax = Inf
    ),
    fill = "gray100",
    alpha = 0.01
  ) +  
  labs(x = "Date",
       y = "New Cases per Day") +
  geom_vline(xintercept = as.Date("2020-03-20"), linetype = 4) +
 scale_y_continuous(breaks = seq(0, 100, by=20),
                        limits = c(0, 100)) +  
  geom_label_repel(
    data = filter(nysCounty2, t == max(t)),
    aes(
      x = t,
      y = I,
      label = sprintf("%i", I),
      fill="red"
    ),
    show.legend = FALSE,
    segment.color = "black",
    nudge_x = -10,
    nudge_y = 5
  ) +
  facet_wrap( ~ county, ncol = 4) +
  theme(
    plot.background = element_rect(fill = "linen"),
    panel.background = element_rect(fill = "linen"),
    strip.background = element_rect(fill = "grey80"),
    strip.text = element_text(color = "black"),
    legend.position = "bottom",
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      size = 1
    ),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
    plot.subtitle = element_text()         
  ) +
  ggtitle(label = "New Cases by County", subtitle= "Counties with a maximum of between 10 and 100 New Cases on a single day")
```

```{r County New 3,  fig.height=20, fig.width=20}
ggplot(data = nysCounty3) +
     geom_point(mapping = aes(x = t, y = I)) +
     geom_point(data = filter(nysCounty3, t == max(t)),
             mapping = aes(x = t, y = I),
             color = "red") +
  geom_rect(
    aes(
      xmin = as.Date("2020-03-20"),
      xmax = as.Date("2020-04-03"),
      ymin = 0,
      ymax = Inf
    ),
    fill = "gray100",
    alpha = 0.01
  ) +  
     labs(x = "Date",
          y = "New Cases per Day") +
     scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10),
                        limits = c(0, 10)) +
     geom_vline(xintercept = as.Date("2020-03-20"), linetype = 4) +
     geom_label_repel(
          data = filter(nysCounty3, t == max(t)),
          aes(
               x = t,
               y = I,
               label = sprintf("%i", I),
               fill="red"
          ),
          show.legend = FALSE,
          segment.color = "black",
          nudge_x = -10,
          nudge_y = 2
     ) +
     facet_wrap(~ county, ncol = 4) +
     theme(
          plot.background = element_rect(fill = "linen"),
          panel.background = element_rect(fill = "linen"),
          strip.background = element_rect(fill = "grey80"),
          strip.text = element_text(color = "black"),
          legend.position = "bottom",
          panel.border = element_rect(
               colour = "black",
               fill = NA,
               size = 1
          ),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()              
     ) +
     ggtitle(label = "New Cases by County", subtitle = "Counties with a maximum of less than 10 New Cases on a single day")
```

#  Cummulative Cases by County

```{r County Cummualtive, fig.width=7.5, fig.height=10}
nys %>% 
  inner_join(NYpop, by=c("f" = "GEOID")) %>% 
  group_by(county) %>% 
  filter(t == max(t)) %>% 
  ggplot(aes(x=C/estimate, y=county, size=estimate, color=estimate)) +
  geom_point() + 
  annotate(geom = "rect", 
           xmin = c(0, 0.01, .02, .03, .04),
           xmax = c(0.01, .02, .03, .04, .05),
           ymin = -Inf,
           ymax = Inf,
           fill = c(o_i_colors[1:5]),
           alpha = 0.1) +
  scale_x_continuous(labels = percent) +
  scale_y_discrete(limits = sort(nys$county, decreasing = TRUE))  +
  scale_size(labels = comma) + 
  scale_color_viridis_c(labels = comma) +
  labs(x = "Percentage of Population Infected", y=NULL) +
  guides(color = guide_legend(title = "Population"), 
         size = guide_legend(title = "Population")) +
  ggtitle(label = "Total Confirmed Cases by County", subtitle="Percent of Population") + 
  theme(
          plot.background = element_rect(fill = "linen"),
          panel.background = element_rect(fill = "linen"),
          strip.background = element_rect(fill = "linen"),
          strip.text = element_text(color = "black"),
          legend.background = element_rect(fill="linen"),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()            
     ) 

#ggplot(data = nys) +
#     geom_point(mapping = aes(x = t, y = C)) +
#     geom_point(data = filter(nys, t == max(t)),
##             mapping = aes(x = t, y = C),
#             color = "red") +
#  geom_rect(
#    aes(
#      xmin = as.Date("2020-03-20"),
#      xmax = as.Date("2020-04-03"),
#      ymin = 0,
#      ymax = Inf
#    ),
#    fill = "gray100",
#    alpha = 0.01
#  ) +  
#     geom_label_repel(
#          data = filter(nys, t == max(t)),
#          aes(
#               x = t,
#               y = C,
#               label = sprintf("%i", C),
#               fill="red"
#          ),
#          show.legend = FALSE,
#          segment.color = "black",
#          nudge_x = -10,
#          nudge_y = -1
#     ) +
#     labs(x = "Date",
#          y = "Log Scale of Cumulative Cases") +
#     scale_y_log10(breaks = c(1, 10, 100, 1000, 10000)) +
#     geom_vline(xintercept = as.Date("2020-03-20"), linetype = 4) +
#     facet_wrap(~ county, ncol = 4) +
#     ggtitle(label = "New Cases by County", subtitle="Cummulative") +
#     theme(
#          plot.background = element_rect(fill = "linen"),
#          panel.background = element_rect(fill = "linen"),
#          strip.background = element_rect(fill = "grey80"),
#          strip.text = element_text(color = "black"),
#          legend.position = "bottom",
#          panel.border = element_rect(
#               colour = "black",
#               fill = NA,
#               size = 1
#          ),
#          panel.grid.major = element_blank(),
#          panel.grid.minor = element_blank(),
#          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
#          plot.subtitle = element_text()            
#     ) 
     
```

# Percentage of Tests performed that were Positive by County

```{r County Percentage, fig.height=50, fig.width=20}
ggplot(data = nys) +
     geom_point(mapping = aes(x = t, y = I * 100 / T)) +
     geom_point(data = filter(nys, t == max(t)),
             mapping = aes(x = t, y = I*100/T),
             color = "red") +
  geom_rect(
    aes(
      xmin = as.Date("2020-03-20"),
      xmax = as.Date("2020-04-03"),
      ymin = 0,
      ymax = Inf
    ),
    fill = "gray100",
    alpha = 0.01
  ) +    
     geom_smooth(mapping = aes(x = t, y = I * 100 / T)) +
     geom_label_repel(
          data = filter(nys, t == max(t)),
          aes(
               x = t,
               y = I * 100 / T,
               label = sprintf("%.1f %%", I * 100 / T),
               fill="red"
          ),
          show.legend = FALSE,
          segment.color = "black",
          nudge_x = -10,
          nudge_y = 50
     ) +
     geom_ribbon(aes(ymin = 0,
                     ymax = T / 60,
                     x = t),
                 alpha = 0.3,
                 fill = "tomato") +
     scale_y_continuous(sec.axis = sec_axis(~ . * 60, name = "Total Tests Performed")) +
     geom_vline(xintercept = as.Date("2020-03-20"), linetype = 4) +
     geom_hline(yintercept = 0) +
     labs(x = "Date",
          y = "Percentage of Tests Performed that were Positive") +
     facet_wrap(~ county, ncol = 4) +
     ggtitle(label = "Percentag Positive by County", subtitle = "Number of Tests Performed") +
     theme(
          plot.background = element_rect(fill = "linen"),
          panel.background = element_rect(fill = "linen"),
          strip.background = element_rect(fill = "grey80"),
          strip.text = element_text(color = "black"),
          legend.position = "bottom",
          axis.text.y.right = element_text(color = "tomato"),
          axis.title.y.right = element_text(color = "tomato"),
          panel.border = element_rect(
               colour = "black",
               fill = NA,
               size = 1
          ),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),        
          plot.subtitle = element_text(color = "tomato")
     )
```
\pagebreak

# Deaths by County. 

<p class="intro">New York City deaths are combined from New York, Bronx, Kings, Queens and Richmond counties.</p>

```{r County Deaths 1, fig.height=5, fig.width=20}
ggplot(data = NYCountyDeaths1) +
     geom_point(mapping = aes(x = date, y = d)) +
     geom_point(data = filter(NYCountyDeaths1, date == max(date)),
             mapping = aes(x = date, y = d),
             color = "red") +
     geom_label_repel(
          data = filter(NYCountyDeaths1, date == max(date)),
          aes(
               x = date,
               y = d,
               label = sprintf("%i", d),
               fill="red"
          ),
          show.legend = FALSE,
          segment.color = "black",
          nudge_x = -10,
          nudge_y = 10
     ) +
     labs(x = "Date",
          y = "Deaths") +
     ylim(0, 1000) +
     facet_wrap( ~ county, ncol = 4) +
     theme(
          plot.background = element_rect(fill = "linen"),
          panel.background = element_rect(fill = "linen"),
          strip.background = element_rect(fill = "grey80"),
          strip.text = element_text(color = "black"),
          legend.position = "bottom",
          panel.border = element_rect(
               colour = "black",
               fill = NA,
               size = 1
          ),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()     
     ) +
     ggtitle(label = "Deaths by County", subtitle="Counties with a maximum of over 100 deaths on a single day")
```

```{r County Deaths 2, fig.height=10, fig.width=20}
ggplot(data = NYCountyDeaths2) +
     geom_point(mapping = aes(x = date, y = d)) +
     geom_point(data = filter(NYCountyDeaths2, date == max(date)),
             mapping = aes(x = date, y = d),
             color = "red") +
     geom_label_repel(
          data = filter(NYCountyDeaths2, date == max(date)),
          aes(
               x = date,
               y = d,
               label = sprintf("%i", d),
               fill="red"
          ),
          show.legend = FALSE,
          segment.color = "black",
          nudge_x = -10,
          nudge_y = 2
     ) +
     labs(x = "Date",
          y = "Deaths") +
     ylim(0, 100) +
     facet_wrap( ~ county, ncol = 4) +
     theme(
          plot.background = element_rect(fill = "linen"),
          panel.background = element_rect(fill = "linen"),
          strip.background = element_rect(fill = "grey80"),
          strip.text = element_text(color = "black"),
          legend.position = "bottom",
          panel.border = element_rect(
               colour = "black",
               fill = NA,
               size = 1
          ),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()             
     ) +
     ggtitle(label = "Death by County", subtitle="Counties with a maximum between 10 and 100 deaths on a single day")
```

```{r County Deaths 3, fig.height=30, fig.width=20}
ggplot(data = NYCountyDeaths3) +
     geom_point(mapping = aes(x = date, y = d)) +
     geom_point(data = filter(NYCountyDeaths3, date == max(date)),
             mapping = aes(x = date, y = d),
             color = "red") +
     labs(x = "Date",
          y = "Deaths") +
     scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10),
                        limits = c(0, 10)) +
     geom_label_repel(
          data = filter(NYCountyDeaths3, date == max(date)),
          aes(
               x = date,
               y = d,
               label = sprintf("%i", d),
               fill="red"
          ),
          show.legend = FALSE,
          segment.color = "black",
          nudge_x = -10,
          nudge_y = 2
     ) +
     facet_wrap( ~ county, ncol = 4) +
     theme(
          plot.background = element_rect(fill = "linen"),
          panel.background = element_rect(fill = "linen"),
          strip.background = element_rect(fill = "linen"),
          strip.text = element_text(color = "black"),
          legend.position = "bottom",
          panel.border = element_rect(
               colour = "black",
               fill = NA,
               size = 1
          ),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()            
     ) +
     ggtitle(label="Deaths by County", subtitle = "Counties with a maximum of 10 or fewer deaths on a single day")
```     

# Estimation of R<sub>t</sub> in New York State by County

<p class="intro">Counties with fewer than 11 cases per week, on average, do not have enough data to accurately estimate R<sub>t</sub>.  For those counties, R<sub>t</sub> and the 95% Confidence Interval are listed as NA.  On the map, they are left blank.</p>

```{r Counties R, fig.width=7.5, fig.height=10}
ggplot(ny_map4, aes(y=County, x = R, size = Population, color = Population)) +
     geom_point() +  
     geom_pointrange(aes(xmin = Confl, 
                         xmax = Confh)) +
     annotate(geom = "rect", 
              xmin = c(0, 0.5, 1, 1.2, 1.5),
              xmax = c(0.5, 1, 1.2, 1.5, 2),
              ymin = -Inf,
              ymax = Inf,
              fill = c(o_i_colors[1:5]),
              alpha = 0.1) +
     scale_x_continuous(breaks = c(0.5, 1, 1.2, 1.5, 2)) +
     scale_size(labels = comma,
                breaks = c(1000, 5000, 10000,50000,100000,200000, 500000, 1000000, 2000000, 3000000),
                range = c(0.1,3)) + 
     scale_color_viridis_c(labels = comma,
                breaks = c(1000, 5000, 10000,50000,100000,200000, 500000, 1000000, 2000000, 3000000)) +
     labs(x = bquote(R[t]), y=NULL) +
     guides(color = guide_legend(), 
            size = guide_legend()) +
     scale_y_discrete(limits = sort(ny_map4$County, decreasing = TRUE))  +
     ggtitle(label= bquote(R[t]), subtitle = "Estimation by County") +
     theme(
          rect = element_rect(fill = "linen"),
          panel.background = element_rect(fill = "linen"),
          strip.background = element_rect(fill = "linen"),
          plot.background = element_rect(fill = "linen"),
          legend.background = element_rect(fill = "linen"),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()             
     )    
```

# Map of R<sub>t</sub> in New York State by County

```{r NY Map R, fig.width= 20, fig.height=20}
ggplot(data = ny_map3) +
     geom_sf(aes(fill = R)) +
     scale_fill_stepsn(
          breaks = c(0.5, 1, 1.2, 1.5),
          colours = o_i_colors[2:8]) +
     labs(fill = bquote(R[t])) +
     geom_sf_text(aes(label = County, geometry=geometry)) +
     ylim(40, 46) +
     ggtitle(label = bquote(R[t]), subtitle="Estimation by County") +
     theme(
          rect = element_rect(fill = "linen"),
          plot.background = element_rect(fill = "linen"),
          legend.background = element_rect(fill = "linen"),
          axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()     
     )   
```

# Estimation of R<sub>t</sub> in the United States by State

```{r States R, fig.width=7.8, fig.height=10}
ggplot(us_map4, aes(y=State, x = R, size = Population, color = Population)) +
     geom_point() + 
     geom_pointrange(aes(xmin = Confl, 
                         xmax = Confh)) +
     annotate(geom = "rect", 
              xmin = c(0, 0.5, 1, 1.2, 1.5),
              xmax = c(0.5, 1, 1.2, 1.5, 2),
              ymin = -Inf,
              ymax = Inf,
              fill = c(o_i_colors[1:5]),
              alpha = 0.1) +
     scale_x_continuous(breaks = c(0.5, 1, 1.2, 1.5, 2)) +
     scale_size(labels = comma,
                breaks = c(1000000, 5000000, 10000000,20000000,30000000,40000000, 50000000),
                range = c(0.1,3)) + 
     scale_color_viridis_c(labels = comma,
                breaks = c(1000000, 5000000, 10000000,20000000,30000000,40000000, 50000000)) +  
     labs(x= bquote(R[t]), y=NULL) +
     guides(color = guide_legend(), 
            size = guide_legend()) +
     scale_y_discrete(limits = sort(us_map4$State, decreasing = TRUE)) +
     ggtitle(label = bquote(R[t]), subtitle = "Estimate by State") +
     theme(
          panel.background = element_rect(fill = "linen"),
          rect = element_rect(fill = "linen"),
          plot.background = element_rect(fill = "linen"),
          legend.background = element_rect(fill = "linen"),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()             
     )    
     
```

# Map of R<sub>t</sub> in the United States

```{r US Map R, fig.width= 20, fig.height=8.4}
ggplot(data = us_map48) +
     geom_sf(aes(fill = R, 
                 geometry = geometry)) +
     scale_fill_stepsn(
       breaks = c(0.5, 1, 1.2, 1.5),
       colours = o_i_colors[2:8] ) +
     geom_sf_text(aes(label = STUSPS, geometry=geometry)) +
     labs(fill = bquote(R[t])) +
     ggtitle(label =  bquote(R[t]), subtitle = "Estimation by State") +
     theme(
          rect = element_rect(fill = "linen"),
          plot.background = element_rect(fill = "linen"),
          legend.background = element_rect(fill = "linen"),
          axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          plot.title = element_text(face="bold", size=rel(1.5), hjust=0),
          plot.subtitle = element_text()    
     )     
```

\pagebreak

<footer>
  <h2 id="footnote-label">Footnotes</h2>
  <ol>
     <li id=#D1>New York State Department of Health data: https://health.data.ny.gov/resource/xdss-u53e.json </li>
     <li id=#D2>New York Times data: https://github.com/nytimes/covid-19-data </li>
     <li id="R">American Journal of Epidemiology
Vol. 178, No. 9 September 15, 2013
A New Framework and Software to Estimate Time-Varying ReproductionNumbers During Epidemics
  <a href="#R-ref" aria-label="Back to content">+</a></li>
      <li id="SI">Emerging Infectious Disease
Volume 26, Number 6June 2020
Research Letter
Serial Interval of COVID-19 among Publicly Reported Confirmed Cases 
  <a href="#SI-ref" aria-label="Back to content">+</a></li>
  </ol>
</footer>